Parameter sensitivity and identifiability for a biogeochemical model of hypoxia in the northern {G}ulf of {M}exico\tnoteref{mytitlenote}

### 2017-03-26 ident_manu.Rnw
methods for calibration/extrapolation written

% R dependencies
<<include = F, cache = F>>=
source('R/funcs.r')
library(tidyr)
library(ggplot2)
library(dplyr)
library(gridExtra)
library(Hmisc)
library(ggrepel)
library(english)
library(purrr)
library(grid)
library(lubridate)
# library(raster)
library(rgeos)
library(ggpolypath)
library(readxl)
library(ggmap)
@

@@

A local sensitivity analysis was performed by evaluating the change in state variables following perturbation of each parameter from its original value \citep{Soetaert10,RDCT16}. Parameters were individually perturbed by 50\% of the original values and sensitivity $S$ was estimated for each time step $i$ given a change for parameter $j$ as:

@@

Parameter selection for model calibration must consider the competing objectives of increased precision with parameter inclusion and reduced identifability as it relates to optimization.  An additional challenge is a large number of combinations of parameter sets, which complicates selection given sensitivity differences and desired ecological categories of each parameter (e.g., practitioners may only be interested in optics parameters).  \cref{fig:combnex} provides a simple graphic of the unique number of combinations that are possible for different subsets of `complete' parameter sets of different sizes (i.e., $n$ choose $k$ combinations, $n!/\left(k!\left(n-k\right)!\right)$).  The number of unique combinations increases with the total parameters in the set and is also maximized for moderate selections (e.g., selecting half the total).  For example, over 10$^{14}$ combinations are possible by selecting 25 parameters from a set of 50.

@@

\subsection{Model calibration with parameter subsets}

% we want to show that the results have relevance for the larger three-dimensional model and that model output for calibration to different parameters sets will have differing precision given 1) differences in sensitivity to included ina  subsets and 2) differences given proportion of parameters selected from different categories.

% we use a time series from Weeks Bay NERRS because data are continuous, initial conditions were modified to match weeks bay conditions.  DO time series was filtered to remove tidal advection component and air/sea gas exhange was used.  This is not a perfect example but works for our purposes to demonstrate differeing precision as directly related to identifiability of subsets and paramter types in a subset. Thus we were primarily concerned with showing differences in calibration results between paramters sets and not an absolute description of the time series.    
% note that we are not concerned with the magnitude of the error but the differences in error between parameter subsets and number of runs to convergence.

@@

@@

\bibliography{refsshr}
\clearpage

@@

p5 <- ggmap(map) + 
  geom_polypath(data = toplo5, aes(x = long, y = lat, group = grp), alpha = 0.5) +
  geom_path(data = toplo6, aes(x = long, y = lat, group = grp, colour = parm), alpha = 0.9, size = 1) + 
  scale_y_continuous(expand = c(0, 0)) + 
  scale_x_continuous(expand = c(0, 0)) +
  scale_colour_manual(values = cols, labels = labs) +
  coord_equal() + 
  mytheme %+replace%
  theme(legend.position = 'none') +
  ggtitle(label = NULL, subtitle = '(e) Estimated areal extents on August 29')

@@

ylab <- expression(paste('Hypoxic area ( ', km^2, ' x 1000)'))
grid.arrange(
  left = textGrob(ylab, rot = 90),
  arrangeGrob(
    arrangeGrob(pA, pC, ncol = 2),
    pleg, 
    arrangeGrob(pB, pD, ncol = 2), 
    pE,
    ncol = 1, 
    heights = c(1, 0.1, 1, 1.4)
  )
)

>>>>

% R dependencies
<<include = F, cache = F>>=
source('R/funcs.r')
library(tidyr)
library(ggplot2)
library(dplyr)
library(gridExtra)
library(Hmisc)
library(ggrepel)
library(english)
library(purrr)
library(grid)
library(lubridate)
# library(raster)
library(rgeos)
library(ggpolypath)
library(readxl)
library(ggmap)
library(ggsn)
@

@@

A local sensitivity analysis was performed by evaluating the change in state variables following perturbation of each parameter from its original value \citep{Soetaert10,RDCT17}. Parameters were individually perturbed by 50\% of the original values and sensitivity $S$ was estimated for each time step $i$ given a change for parameter $j$ as:

@@

Parameter selection for model calibration must consider the competing objectives of increased precision with parameter inclusion and reduced identifability as it relates to optimization.  An additional challenge is a large number of combinations of parameter sets, which complicates selection given sensitivity differences and desired ecological categories of each parameter (e.g., parameters for a group of related structural equations could be of interest).  \cref{fig:combnex} provides a simple graphic of the unique number of combinations that are possible for different subsets of `complete' parameter sets of different sizes (i.e., $n$ choose $k$ combinations, $n!/\left(k!\left(n-k\right)!\right)$).  The number of unique combinations increases with the total parameters in the set and is also maximized for moderate selections (e.g., selecting half the total).  For example, over 10$^{14}$ combinations are possible by selecting 25 parameters from a set of 50.

@@

\subsection{Model calibration and extrapolation}

Selected parameter subsets from the seven heuristics were used to calibrate the \ac{zerod} model.  This analysis demonstrated that different parameter sets can produce model output with differing precision and calibration effort given 1) differences in variable sensitivity to parameters in each subset, 2) differences in the parameters selected from different categories (heuristic difference), and 3) differences in identifiability values.  Generally, this analysis was a proof of concept that the selection heuristics can lead to an `optimal' model by maximizing the balance between sensitivity and identifiability.

A time series of \ac{do} from Weeks Bay, Alabama was used for model calibration.  Daily observations from January 1\textsuperscript{st},  2006 to January 1\textsuperscript{st}, 2007 at the Weeks Bay station of the Weeks Bay National Estuarine Reserve were used \citep{Wenner04}.  We chose this location because of the availability of data for model calibration compared to bottom waters in the northern \ac{gom} and the system exhibits hypoxia from eutrophication.  This application also provided a practical example of how the FishTank model could be calibrated using data from alternative locations.  Initial conditions for model input were modified to match conditions in Weeks Bay using observed water quality and nutrient data (all data from \url{http://cdmo.baruch.sc.edu/}).  The \ac{do} time series was filtered to reduce physical effects of the tidal advection component on the observed data (\textit{sensu} \citealt{Beck15b}).  As such, the filtered data were comparable to a time series with no physical advection to match conditions of the \ac{zerod} FishTank model with no hydrodynamic component.

Each parameter subset was calibrated to most closely match the model output to the observed time series.  Parameter values were searched using the optim function in R that implemented the limited memory modification of the BFGS quasi-Newton method \citep{Byrd95,Nocedal06,RDCT17}.  This opimization scheme searches for values within the user-defined lower and upper bounds of each parameter.  We set the algorithm to search within +/- 50\% of the default parameter values.  While we recognize that `actual' parameter values could possibly be outside of this range, constraining the values to those tested herein was expected to produce output that was more interpretable within the context of the sensitivity functions defined above.  Moreover, precise limits on each parameter were less of a concern for the current analysis as our primary goal was to demonstrate differences in precision as a function of sensitivity and identifiability.  The optimal values for each parameter were identified based on convergence of the model results as a minimization of the root-mean squared error between the observed and modelled data.

As a final analysis, effects of parameter changes on the larger 3-dimensional model were evaluated to demonstrate that results from the \ac{zerod} FishTank model can be extrapolated.  For simplicity, only the most sensitive parameter in each category was selected.  The \ac{cgem} model was run for one year using the default parameter conditions and by increasing each of the selected parameters by 50\%.  This produced six spatially-referenced time series from which the changes in the daily hypoxic area (number of cells $<$ 64 mmol \ac{do}) were estimated in the nothern \ac{gom}.

@@

\subsection{Model calibration and extrapolation}

Calibration of parameter values for each of the seven selection heuristics produced model output that varied in precision and computational effort to identify optimal values.  The seven evaluated subsets were those for \ac{do} in each of the four categories of parameters in \cref{tab:heurist1}, including the single parameter for temperature.  The subsets in \cref{tab:heurist2,tab:heurist3} for the parameters selected independent of category and equally within each category were also evaluated.
<<>>=
NULL
@

% starting rmse of model without calibration
\cref{fig:areachg}
% does overall area change correlate with degree of sensitivity?

@@

\bibliography{refs}
% \bibliography{refsshr}
\clearpage

@@

p5 <- ggmap(map) + 
  geom_polypath(data = toplo5, aes(x = long, y = lat, group = grp), alpha = 0.5) +
  geom_path(data = toplo6, aes(x = long, y = lat, group = grp, colour = parm), alpha = 0.8, size = 0.8) + 
  scale_y_continuous(expand = c(0, 0)) + 
  scale_x_continuous(expand = c(0, 0)) +
  scale_colour_manual(values = cols, labels = labs) +
  coord_equal() + 
  mytheme %+replace%
  theme(legend.position = 'none') +
  ggtitle(label = NULL, subtitle = '(e) Estimated areal extents on August 29, default extent in grey') + 
  scalebar(location = 'bottomleft', y.min = 28.35, y.max = 28.8, x.min = -93.3, x.max = -92, 
    dist = 20, dd2km= TRUE, model='WGS84', st.dist=.1, st.size = 2, height = 0.08)

@@

ylab <- expression(paste('                                               Hypoxic area ( ', km^2, ' x 1000)'))
grid.arrange(
  arrangeGrob(
    arrangeGrob(pA, pC, ncol = 2),
    pleg, 
    arrangeGrob(pB, pD, ncol = 2), 
    pE,
    ncol = 1, 
      left = textGrob(ylab, rot = 90),
    heights = c(1, 0.1, 1, 1.4)
  )
)

### 2017-03-26 ident_manu.tex
methods for calibration/extrapolation written

\title{Parameter sensitivity and identifiability for a biogeochemical model of hypoxia in the northern {G}ulf of {M}exico\tnoteref{mytitlenote}}
\tnotetext[mytitlenote]{Version: Sun Mar 12 12:53:46 2017 -0500, \href{https://github.com/fawda123/identifiability/commit/39a5c73735c94c696b2422eddee88342245006dd}{39a5c73735c94c696b2422eddee88342245006dd}}

@@

>>>>

\title{Parameter sensitivity and identifiability for a biogeochemical model of hypoxia in the northern {G}ulf of {M}exico\tnoteref{mytitlenote}}
\tnotetext[mytitlenote]{Version: Fri Mar 17 14:00:58 2017 -0500, \href{https://github.com/fawda123/identifiability/commit/4e88d0837d798238902ff4eb6348b72e9143c557}{4e88d0837d798238902ff4eb6348b72e9143c557}}

@@

\subsection{Model calibration and extrapolation}

Calibration of parameter values for each of the seven selection heuristics produced model output that varied in precision and computational effort to identify optimal values.  The seven evaluated subsets were those for \ac{do} in each of the four categories of parameters in \cref{tab:heurist1}, including the single parameter for temperature.  The subsets in \cref{tab:heurist2,tab:heurist3} for the parameters selected independent of category and equally within each category were also evaluated.
NULL

% starting rmse of model without calibration
\cref{fig:areachg}
% does overall area change correlate with degree of sensitivity?
